The effect of symbols in government communications on citizens' trust, and its interaction with substantive information

### 2018-07-21 paper2_information and symbols_02.Rmd
Add files via upload

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements can increase citizens' trust via two mechanisms: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by leading citizens to dedicate less attention to the substantive information in the communication, and thus causing them to disregard the content of logically unpersuasive information that would have otherwise decreased their trust. I examine these expectations via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was stronger among those who were exposed to the logically unpersuasive policy plans, yet distracted by the symbols and thus less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their underestimation of public-service performance. One of the main remedies which has been proposed for this problem is greater investment in direct government communications with citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@deacon_2001; @eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant public administration research on the effect of government public communications on citizens attitudes has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on organizations' performance and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

Recently, a few studies pointed to the additional effect of *symbolic* elements that are entangled in government communications, such as  agencies' names and brand logos, figures, images and celebrity endorsements [@alonbarkat_2017; @karens_2016; @teodoro_2018]. These studies demonstrated that government organizations may cause citizens to view them more favorably through familiar and well-designed symbolic elements that evoke positive associations and emotions. Still, we lack a theory and robust empirical evidence of the exact mechanisms through which these symbols increase citizens trust. Additionally, these previous studies did not examine the interaction between the symbolic elements and the substantive content of information in public communications.

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a democratic theory standpoint, the effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands in tension with the normative expectation that citizens' judgments should be based on their critical thinking. Mainly, we should be concerned about the possibility that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in governemnt communications affect citizens' trust in government organizations and their policies?* Accordingly, *does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for the varying psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust via two mechanisms: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard the content of logically unpersuasive information that would have otherwise decreased their trust.

@@

I find that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans, yet less inclined to think about the policy content due to distraction by the symbols. I discuss the theoretical and normative implications of these findings.

@@

Symbols are fundamental for almost any form of human interaction, as acknowledged by early scholars of social science [@blumer_1986]. A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. My focus is on those symbols that are strategically used by government organizations in their public communications, are designed to distinguish them and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can be also found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

By comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the names of US federal agencies not only represent information about their activities, but may also carry specific symbolic meaning and affect that can shape citizens' support for government policies (depending on citizens' political party identification). These two studies provide examples for the role of symbols in shaping citizens' attitudes. Still, they did not effectively disentangle the effects of the symbolic and informational aspects of the studied public communications.

@@

According to ELM, the responses of individuals to persuasive communications depends on the amount of thinking that they dedicate to the message content. Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little amount of cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion through the central route is expected to be more enduring, resistant and predictive of behavior, compared with peripheral route persuasion. Of course, much of the time, people are located in between the two ends of the continuum, and thus persuasion is determined by a mixture of these processes.

@@

ELM enables to disentangle the distinct effects of different elements in communications. Some elements are processed differently by people at different locations along the elaboration continuum. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as argumentation, and systematically scrutinize it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all. For example, the physical attractiveness of the source normally does not represent any substantive argument, yet it can serve as a simple cue. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]. The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuades people directly through the peripheral route, yet at the same time they can also act as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

[^footnote_multifunction_factors]: @petty_1986 have suggested that we are more likely to find complex relations of multiple mechanisms in those situations where people are located in the middle of the elaboration continuum. When people are neither extremely motivated or unmotivated to think, they may rely on different elements in the communications to decide if the message is worth or needs considering.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust through the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).[^footnote_interaction_strong_information]

Symbolic elements may decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and absorbing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message - government organizations and policies - is likely to yield negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood. The latter expectation entails an interaction between symbols and the substantive information in the communications.

@@

In sum, I expect that:

*H<sub>1</sub> - Symbolic elements in public organizations' communications are likely to increase citizens' trust in the organizations and in the policies presented in the communications.*

*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that would have decreased citizens' trust.*

@@

*H<sub>4</sub> - The abovementioned moderated effect of symbolic elements on citizens' trust H<sub>2</sub> is likely to be mediated by their decreasing effect on citizens’ amount of thinking H<sub>3</sub> (a moderated mediation hypothesis).*

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans. Such communications can be found in publicized reports, pamphlets, websites, new media pages and ads. Focusing on communication of future policy plans, as opposed to communications on performance of existing policies, enables me to utilize a real-world empirical case without compromising the reliability of the manipulation of the content. This is because most citizens do not have prior knowledge of organizations' future policy plans, and thus they will be more willing to trust a fictional communication of an organization with regard to its future policy.

@@

[^footnote_relevance_paper]: The methodological aspects which are relevant for examining the additional hypotheses (i.e. the interaction between symbols with personal relevance) are described in a separate paper. An English translation of the full survey is available from the supplementary appendix.

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM), a national level bureau responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM makes a good case for this study due to its significant investment, over the past years, in public communications. In these various communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (image a. in figure 2). This logo consists a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element on which I focused is the green color, which is strongly associated with taking care of the environment. The EPM's logo and the green color can be seen in almost all of the visual communications of the EPM, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, that previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (image b.) was launched in 2010, and included a series of advertisements in which Friedman was presented with a green grass head, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (image c.) was launched in 2017, and was focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

In the course of the survey experiment, I confirmed that the EPA logo and the images of the celebrities from the campaigns are familiar to the participants. `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize EPA logo to some extent and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`% , respectively). In robust analyses I further exclude those who do not recognize these symbols.[^footnote_prior_studies]

@@

[^footnote_prior_studies]: In two prior online survey tests, which I conducted before the experiment, I collected further information about the affect and symbolic associations of these symbols. These results, which are reported in the supplementary appendix, substantiate my empirical assumption that these elements evoke positive emotions Israeli citizens, and that citizens automatically associate them with EPM and its policies. These results also indicate that the images of the celebrity comedians from EPM campaigns are perceived as humoristic, and accordingly, they are expected to position people in a positive mood.

<br>
**Figure 2: symbolic elements of EPM**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/EPM_symbols.png) 
<font size="0.5">

Notes:

a. EPM logo.
b. Tal Friedman, comedian,  EPM campaign “starting to think green” (2010): "Isn't money growing on trees? Moving to environmental consumption and saving at least 6,000 NIS per year!", "starting to think green", "The Environmental Protection Ministry".
c. Ido Rosenblum, comedian, EPM campaign  “taking every bag seriously” (2017): "The reusable bag is again left at home? A tip from Ido: Hang it on the door handle and you will not leave without it", "Taking every bag seriously", "The Environmental Protection Ministry".

@@

The survey was presented to the participants as a study about citizens' attitudes regarding environmental policy issues. The participants were first asked to respond to pre-manipulation questions about their political ideology and their perceptions of the pubic sector.[^footnote_relevance_manipulation] Thereafter, they were told that they will be presented with two government policy plans regarding specific environmental issues, obtained from EPM's official publication of its annual work plan for the year 2018. The policy plans regarded reducing air-pollution in the Bay of Haifa (an area in the north of Israel) and reducing domestic waste and increasing recycling. The two policy plans were presented to each subject in a random order. Each policy plan included the title of the policy goal (i.e. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

@@

Per each policy, subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). Each subject, moreover, was exposed to one strong and one weak policy communication. For example, a subject who first received a strong air-pollution policy, later received the weak recycling policy, and vice versa. The actions which were mentioned in the strong policy plans condition were taken from the real EPM work plan and from other reports and official sources of EPM. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures that are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were decreasing the supervision of polluting factories, and raising citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is displayed in **Appendix A**.

>>>>

*Effective public communications have been proposed as a remedy for citizens' distrust in government. Those studies that have explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the Elaboration Likelihood Model (ELM) from social psychology I hypothesize that symbolic elements can increase citizens' trust in two ways: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by conducing citizens to pay less attention to the substantive information in the communication. As a result citizens disregard logically unpersuasive information that could decrease their trust. I examine these expectations via a survey experiment, focusing on communications of the Israeli Environment Protection Ministry about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust. The effect was stronger among those who were exposed to the logically unpersuasive policy plans. The latter were also distracted by the symbols and thus less inclined to scrutinize their content. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

Public administration scholars are highly concerned about the implications of citizens’ low trust in government, and their underestimation of public-service performance. One of the main remedies proposed for this is greater investment in direct government communications to citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@deacon_2001; @eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant public administration research on the effect of government public communications on citizens attitudes has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2017a; @porumbescu_2017b; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. Public communications are expected to increase citizens' trust insofar as they provide citizens with favorable information on the performance of organizations and logically persuasive explanations for their policies. At the same time, experiment-based studies that explored the responses of citizens to substantive information revealed cognitive biases that distort their processing of information in communications, and undermine its effectiveness [@baekgaard_2016; @grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @hvidman_2016; @james_2011; @james_2014; @james_2017; @marvel_2015; @olsen_2017; @james_2017a; @grimmelikhuijsen_2017].

Recently, a few studies pointed to the additional effect of *symbolic* elements that are entangled in government communications, such as  agencies' names and brand logos, figures, images and celebrity endorsements [@alonbarkat_2017; @karens_2016; @teodoro_2018]. These studies demonstrated that government organizations may cause citizens to view them more favorably through familiar and well-designed symbolic elements that evoke positive associations and emotions. Still, we lack a theory and robust empirical evidence of the exact mechanisms through which these symbols increase citizens’ trust. These previous studies also did not examine the interaction between the symbolic elements and the substantive content of information in public communications.

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a democratic theory standpoint, the effect of symbolic elements can be problematic. One can argue that shaping citizens' perceptions of governments by evoking associations and emotions stands against the normative expectation that citizens' judgments should be based on their critical thinking. We need to be aware that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in government communications affect citizens' trust in government organizations and their policies? Does the effect of symbolic elements depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for various psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change  [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust in two ways: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard the content of logically unpersuasive information that could otherwise decreased their trust.

@@

I found that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans. They were also less inclined to think about the policy content because they were distracted by the symbols. I discuss the theoretical and normative implications of these findings.

@@

Symbols are fundamental for almost any form of human interaction, as acknowledged by early scholars of social science [@blumer_1986]. A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. My focus is on those symbols that are strategically used by government organizations in their public communications, are designed to distinguish them and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can also be found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

By comparison with the abundant research on citizens’ responses to informative elements in public communications [for reviews see: @cucciniello_2016; @grimmelikhuijsen_2017; @james_2017], public administration literature has devoted relatively little attention to the effect of symbolic elements. Recently, however, a few experimental studies tackled the issue. @marvel_2015 analyzed US citizens' responses to a television advertisement of the public Postal Service, which involved various symbolic elements, including the organization's logo, workers' images and background music. His study shows that participants who watched the advertisement evaluated the organization's performance more positively. @teodoro_2018 suggested that the names of US federal agencies not only represent information about their activities, but may also carry specific symbolic meaning and positive/negative "affect" that can shape citizens' support for government policies (depending on citizens' political party identification). These two studies provide examples for the role of symbols in shaping citizens' attitudes. However, they did not effectively disentangle the effects of the symbolic and informational aspects of the public communications studied.

@@

According to ELM, the responses of individuals to persuasive communications depends on how much effort they devote to think about the content of the message (or "the amount of thinking" according to ELM terminology). Persuasion can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these different persuasion mechanisms are likely to cause divergent types of attitudinal change. Persuasion via the central route is expected to be more enduring, resistant and predictive of behavior, compared with peripheral route. Of course, often people are located in between the two ends of the continuum, and thus persuasion is determined by a mixture of these processes.

@@

ELM enables us to disentangle the distinct effects of different elements in communications. Some elements are processed differently by people at different locations along the elaboration continuum. Consider for instance a text that contains a series of logical arguments supporting a certain position. Those at the high end of the continuum would regard this text as argumentation, systematically scrutinizing it based on its logical persuasive quality (central route processing). Conversely, those at the low end of the continuum are more likely to disregard the content of the text, and instead, rely on the number of lines in the text, for example, as a cue for its strength (peripheral route processing). Some elements provide potential cues that lack any logical quality at all. For example, the physical attractiveness of the source normally does not represent any substantive argument, yet it can serve as a simple cue. These latter elements may affect the attitudes of those at the low end, but are less likely to persuade those at the high end [@petty_1986].

@@

Elements in communications can also function as factors that affect people's motivation and/or ability to think. For example, Background music in commercials can distract people and prevent them from scrutinizing their content [@park_1986]. The use of humor in television shows stimulates positive mood and reduces people's ability and motivation to scrutinize their substantive arguments [@nabi_2007; @young_2008]. Hence, elements can have more than one function, and can shape attitudes through several mechanisms. They can serve both as variables that affect the amount of thinking and as potential peripheral cues or arguments. Continuing with the above-mentioned examples, background music and humor can serve as simple cues that persuade people directly through the peripheral route, yet at the same time they can also act as factors that reduce elaboration. Accordingly, they can affect people both directly (as peripheral cues) and indirectly (by decreasing elaboration and thus attenuating the effect of arguments and enhancing the effect of peripheral cues).[^footnote_multifunction_factors]

[^footnote_multifunction_factors]: @petty_1986 have suggested that we are more likely to find complex relations of multiple mechanisms in those situations where people are located in the middle of the elaboration continuum. When people are neither extremely motivated or unmotivated to think, they may rely on different elements in the communications to decide whether the message is worth or needs considering.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and capable to do so. Conversely, symbolic elements are potential peripheral cues, which can affect citizens' trust via the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive description of their policy plans).[^footnote_interaction_strong_information]

Symbolic elements may decrease citizens' ability and motivation to think about the message in several ways. First, they can cause elaboration disruption, by distracting citizens and capturing their attention and cognitive efforts [@petty_1976; @petty_1986]. Second, they can decrease citizens' motivation and ability to think about the message content by creating a positive affect and placing citizens in a positive mood. As explained above, the postulation that positive mood decreases elaboration has been explained, in part, by a mood maintenance mechanism [@mackie_1989; @bless_1990; @schwarz_1991]. This mechanism is particularly relevant to the context of government communications, in which thinking about the content of the message – government organizations and policies – is likely to yield a negative mood among most people [@marvel_2015; @marvel_2015a; @hvidman_2016]. Therefore, when citizens are exposed to symbolic elements that evoke positive feelings, they may seek to avoid thinking about tedious bureaucratic affairs, which would depress their positive mood. The latter expectation entails an interaction between symbols and the substantive information in the communications.

@@

In summary, I expect that:

*H<sub>1</sub> - Symbolic elements in public organizations' communications are likely to increase citizens' trust in those organizations and in the policies presented in their communications.*

*H<sub>2</sub> - The positive effect of symbolic elements on citizens' trust is likely to be enhanced when the communication includes negative or logically unpersuasive information that could otherwise have decreased citizens' trust.*

@@

*H<sub>4</sub> - The abovementioned moderated effect of symbolic elements on citizens' trust H<sub>2</sub> is likely to be mediated by their negative effect on citizens’ amount of thinking H<sub>3</sub> (a moderated mediation hypothesis).*

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans. Such communications can be found in publicized reports, pamphlets, websites, new media pages and ads. I focus on communication of future policy plans, as opposed to communications on performance of existing policies. This allows me to utilize a real-world empirical case without compromising the reliability of the manipulation of the content. This is because most citizens do not have prior knowledge of organizations' future policy plans, and thus they will be more willing to trust a fictional communication of an organization with regard to its future policy.

@@

[^footnote_relevance_paper]: The methodological aspects which are relevant for examining the additional hypotheses (i.e. the interaction between symbols with personal relevance) are described in a separate paper. An English translation of the full survey is available in the paper's supplementary appendix.

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM) that is responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

Methodologically, EPM is a good candidate for this study due to its significant investment, over the past years, in public communications. In these communications, the EPM used an array of symbolic elements, which became widely recognized by the public. Here, I focused on three prominent symbolic elements used in the EPM's public communications, and displayed below in **Figure 2**. The first symbolic element is the unique brand logo of the EPM (Figure 2a). This logo consists of a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second symbolic element which I focused on, is the color green, which is strongly associated with environmental protection. The EPM logo and the green color are used in almost all of EPM’s visual communications, and they represent key elements of its unique "visual identity" [@vandenbosch_2005]. The third element is celebrity endorsements. Specifically, I selected two high-profile celebrity comedians, *Tal Friedman* and *Ido Rosenblum*, who previously endorsed two salient advertising campaigns of the EPM. The campaign starring the comedian Tal Friedman (Figure 2b) was launched in 2010, and included a series of advertisements in which Friedman was presented with a head covered in green grass, and accordingly the campaign's slogan was: "starting to think green". The campaign starring the comedian Ido Rosenblum (Figure 2c) was launched in 2017, and focused on reducing the use of disposable bags.[^footnote_campaigns] The two campaigns were highly successful in terms of their recognizability and likability, as suggested by their evaluation reports by the Government advertising agency.

In the course of the survey experiment, I was able to confirm that the EPA logo and the images of the celebrities from the campaigns were familiar to the participants. A total of `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize the EPA logo to some extent, and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`% , respectively). In robust analyses I further exclude those who do not recognize these symbols.[^footnote_prior_studies]

@@

[^footnote_prior_studies]: In two prior online survey tests, which I conducted before the experiment, I collected further information about the affect and symbolic associations of these symbols. These results are reported in the supplementary appendix and substantiate my empirical assumption that these elements evoke positive emotions in Israeli citizens, and that citizens automatically associate them with EPM and its policies. These results also indicate that the images of the celebrity comedians from EPM campaigns are perceived as humoristic, and accordingly, they are expected to evoke a positive mood in citizens.

<br>
**Figure 2: Symbolic elements of EPM**
![](D:/UNIVERSITY/R/SVIVA/experiment 2 01-2018/myfigures/EPM_symbols.png) 
<font size="0.5">

*Notes:*

a. EPM logo.
b. Tal Friedman, comedian,  EPM campaign “starting to think green” (2010): "Isn't money growing on trees? Moving to environmental consumption and saving at least 6,000 NIS per year!", "starting to think green", "The Environmental Protection Ministry".
c. Ido Rosenblum, comedian, EPM campaign  “taking every bag seriously” (2017): "Did you for forget your reusable carrier bag at home again? A tip from Ido: Hang it on the door handle and you will not leave without it", "Taking every bag seriously", "The Environmental Protection Ministry".

@@

The survey was presented to the participants as a study about citizens' attitudes regarding environmental policy issues. The participants were first asked to respond to pre-manipulation questions about their political ideology and their perceptions of the public sector.[^footnote_relevance_manipulation] Thereafter, they were told that they would be presented with two government policy plans regarding specific environmental issues, extracted from EPM's official 2018 annual work plan. The policy plans included reducing air-pollution in Haifa bay area in the north of Israel, reducing domestic waste and increasing recycling. The two policy plans were presented to each subject in a random order. Each policy plan included the title of the policy goal (e.g. "reducing the levels of air pollution in the Haifa bay"), followed by short descriptions of two actions to be taken by EPM in order to fulfill this goal (30-50 words each).

@@

Per each policy, subjects were randomly allocated to one of two conditions of substantive information (hereafter: information): strong policy (i.e. logically persuasive policy plan) or weak policy (i.e. logically unpersuasive). Each subject, moreover, was exposed to one strong and one weak policy communication. So for example a subject who first received a strong air-pollution policy plan, later received the weak recycling policy plan, and vice versa. The actions which were mentioned in the condition of the strong policy plans were taken from the real EPM work plan and from other reports and official EPM sources. The activities that were included in the weak policy plans condition were fictional, and were designed to represent measures that are likely to be perceived by subjects as extremely incompatible with achieving the policy goals, yet still sufficiently reliable. For instance, the proposed actions for the air-pollution strong policy involved increasing the supervision of polluting factories, and reducing diesel smoke emissions by vehicles, whereas the actions for the weak policy were to decrease the supervision of polluting factories, and raise citizens' awareness of the Ministry's efforts to improve air quality. The full texts of the weak and strong conditions for the two policy plans is included in **Appendix A**.
